\titleformat{\chapter}[hang]{\Huge\bfseries}{\selectfont \thechapter\hsp\textcolor{red}{\emoji{rocket}}\hsp}{0pt}{\Huge\bfseries}

\chapter{Conclusion}

This chapter summarizes the findings of the project, evaluates the achievement of the business and data mining goals, and outlines limitations and future work directions.

\section{Summary of Results}

This project developed a temporal link prediction system for social networks using EvolveGCN, following the CRISP-DM methodology. The work addressed the challenge of predicting future connections in dynamic social networks and generating synthetic network structures.

\subsection*{Link Prediction Performance}

Two training pipelines were implemented and evaluated: the Original (sequence-based) and Incremental (phase-based) approaches. The key findings are:

\begin{itemize}
	\item The \textbf{Original implementation} achieved superior ranking and classification performance, with MAP of 0.87 and Macro F1 of 0.86 on the best configuration (LR=0.001, NM=1).
	\item The \textbf{Incremental implementation} achieved better discrimination ability with AUC of 0.80, suggesting its potential for threshold-independent ranking tasks.
	\item Negative sampling multiplier significantly impacts model behavior: higher values (NM=2) improve AUC but reduce MAP, indicating a trade-off between decision boundary learning and ranking quality.
\end{itemize}

\subsection*{Synthetic Network Generation}

The application to synthetic data revealed significant challenges:

\begin{itemize}
	\item The Original implementation fails to predict edges between synthetic users, attributing to the extreme dimensional collapse (PC1 explaining 95-100\% variance) that prevents the model from learning discriminative features for the homogeneous synthetic user cluster.
	\item Only the Incremental approach with richer embedding space (PC1=76\%) attempts to connect synthetic users, though with poor structural fidelity.
	\item The generated synthetic network exhibits unrealistic properties: average degree 7$\times$ higher than real networks, near-zero modularity indicating absence of community structure, and extremely low average path length due to over-dense connections.
\end{itemize}

\subsection*{Embedding Space Analysis}

The GCN embedding visualization provided critical insights into model behavior:

\begin{itemize}
	\item \textbf{Oversmoothing}: The Original approach suffers from severe dimensional collapse, where node representations become nearly indistinguishable after multiple GCN layers.
	\item \textbf{Real-Synthetic Separation}: The centroid distance between real and synthetic embeddings correlates with classification performance. Models with higher separation (6.29-6.65) achieve better MAP but fail to generalize to synthetic users.
	\item \textbf{Synthetic Homogeneity}: Synthetic users exhibit 11.23$\times$ lower variance than real users, forming an isolated cluster that the model cannot meaningfully connect to the real network structure.
\end{itemize}

\section{Achievement of Success Criteria}

The business success criteria defined in Section~\ref{sec:business_success} are evaluated as follows:

\begin{enumerate}
	\item \textbf{Development of a temporal link prediction model based on EvolveGCN}: \textcolor{drawgreen}{\textbf{Achieved}}. The EvolveGCN-H architecture was successfully implemented with both training pipelines.

	\item \textbf{Quantitative comparison using standard metrics}: \textcolor{drawgreen}{\textbf{Achieved}}. Comprehensive evaluation with MAP, AUC, Precision, Recall, and F1-score was conducted across multiple configurations.

	\item \textbf{Synthetic network generation}: \textcolor{drawred}{\textbf{Failed}}. The model generates predictions on synthetic data, but the resulting networks lack structural realism due to embedding distribution mismatch and severe oversmoothing.
\end{enumerate}

\section{Limitations}

Several limitations were identified during this project:

\begin{itemize}
	\item \textbf{Embedding Distribution Mismatch}: The synthetic BERT embeddings are too homogeneous (within-group cosine similarity of 0.988) compared to real users (0.872), preventing the model from learning meaningful connections for synthetic nodes.

	\item \textbf{Oversmoothing in GCN}: The dimensional collapse observed in the Original implementation limits the model's ability to capture fine-grained user differences, particularly problematic for the link prediction task.

	\item \textbf{Cumulative Edge Assumption}: The assumption that edges persist once created simplifies the task but does not reflect real-world unfollowing behavior, potentially limiting real-world applicability.

\end{itemize}

\section{Future Work}

Based on the findings, several directions for future research are recommended:

\begin{enumerate}
	\item \textbf{Mitigating Oversmoothing}: Implement techniques such as jumping knowledge networks, APPNP-style propagation, or fewer GCN layers to preserve node-level discrimination.

	\item \textbf{Synthetic Data Improvement}: Generate synthetic posts with more diverse prompts or apply embedding perturbation techniques to increase variance and reduce the isolation of synthetic users.

	\item \textbf{Test other models}: Explore alternative temporal graph models that may better capture dynamic patterns without severe oversmoothing.
\end{enumerate}
