\titleformat{\chapter}[hang]{\Huge\bfseries}{\selectfont \thechapter\hsp\textcolor{red}{\emoji{package}}\hsp}{0pt}{\Huge\bfseries}

\chapter{Data Preparation}

This chapter covers the data preparation phase of the CRISP-DM methodology.

\section{Select Data}

The data selection process focuses on preparing the temporal graph data for link prediction using EvolveGCN. The relevant data consists of two main components:

\begin{itemize}
	\item \textbf{Social Network Graphs}: The edge lists from each snapshot representing the directed follower relationships (X follows Y) between users.
	\item \textbf{Post Content}: The textual content from posts used to derive node features through BERT embeddings.
\end{itemize}

For the incremental learning approach, we use the cumulative structure where each snapshot $S_i$ contains all edges from previous snapshots plus new edges added in that time period. The dataset is organized into six temporal snapshots: 2016-2021, 2022, 2023, 2024, Jan-Jul 2025, and Jul 2025.

\subsection*{Feature Selection}

Node features are derived from pre-computed BERT embeddings of user posts. For each user, the embedding is computed as the average of BERT embeddings across all their posts up to the current snapshot:

\begin{equation}
	\mathbf{h}_u^{(t)} = \frac{1}{|P_u^{(t)}|} \sum_{p \in P_u^{(t)}} \text{BERT}(\text{content}_p)
\end{equation}

where $P_u^{(t)}$ is the set of posts by user $u$ up to snapshot $t$. This yields a 768-dimensional feature vector per node, capturing the semantic content of each user's contributions.

\section{Clean Data}

\subsection*{Missing Values}

The data quality analysis revealed minimal missing values:

\begin{itemize}
	\item The \texttt{language} attribute has 877 missing values (0.48\%) in the aggregated dataset
	\item All other attributes are 100\% complete
	\item No empty post content was found
\end{itemize}

Missing language values are attributed to automatic language detection failures on very short or mixed-language posts. For the link prediction task, this does not affect model performance since features are derived from BERT embeddings rather than language labels.

\subsection*{Duplicate Handling}

Content-level duplicates were identified across snapshots, with the duplicate rate increasing over time (from 1.79\% in 2016-2021 to 10.04\% in Jul 2025). Two types of content duplicates exist:

\begin{itemize}
	\item \textbf{Intra-user duplicates}: The same user posting identical content multiple times (reposts)
	\item \textbf{Inter-user duplicates}: Different users sharing the same post content (shared posts, quotes, or viral content)
\end{itemize}

For inter-user duplicates, the BERT embeddings computed for the identical posts are necessarily identical, since the embedding function is deterministic and content-based. These duplicates do not negatively impact the task since:

\begin{enumerate}
	\item The network structure is defined by edges, not post content
	\item Post IDs remain unique even when content is duplicated
	\item User embeddings aggregate all posts, so duplicate content reinforces rather than distorts the representation
	\item Identical embeddings for shared content naturally capture content-based user similarity
\end{enumerate}

\section{Construct Data}

\subsection*{Node Feature Construction}

Temporal node features are constructed respecting the incremental nature of the dataset. For snapshot $t$, the features for each node are computed using only posts that exist up to and including snapshot $t$:

\begin{equation}
	\mathbf{h}_u^{(t)} = \frac{1}{|P_u^{(t)}|} \sum_{p \in P_u^{(t)}} \text{BERT}(\text{content}_p)
\end{equation}

where $P_u^{(t)} = \{p : \text{snapshot}(p) \leq t\}$.

This temporal feature construction ensures no data leakage: at training time for snapshot $t$, the model only has access to information that would have been available at that point in time.

\subsection*{Graph Construction}

The social network is represented as a directed graph $G_t = (V_t, E_t)$ at each snapshot $t$, where:

\begin{itemize}
	\item $V_t$ is the set of nodes (users) that appear in any edge up to snapshot $t$ (cumulative)
	\item $E_t$ is the set of directed edges (follower relationships) present at snapshot $t$
	\item Edges represent ``X follows Y'' relationships and are treated as unweighted
\end{itemize}

The cumulative assumption states that once an edge exists, it persists in all subsequent snapshots. 
%This reflects the nature of follower relationships on social platforms.

% \subsection{Node ID Remapping}

% Original user IDs from the Gab platform are remapped to contiguous indices $[0, N-1]$ where $N$ is the total number of unique nodes across all snapshots. This remapping:

% \begin{itemize}
% 	\item Enables efficient sparse tensor operations
% 	\item Reduces memory footprint by eliminating large, sparse ID spaces
% 	\item Maintains a bidirectional mapping between original and remapped IDs for interpretability
% \end{itemize}

\section{Integrate Data}

Data integration involves combining the edge lists with post-derived features:

\begin{enumerate}
	\item \textbf{Edge Loading}: All \texttt{.edg} files are loaded and annotated with their corresponding snapshot index
	\item \textbf{Post-to-User Mapping}: A mapping from post ID to user ID enables attribution of posts to nodes
	\item \textbf{Post-to-Snapshot Mapping}: A mapping from post ID to snapshot index enables temporal feature filtering
\end{enumerate}

The integration produces a unified dataset object containing:
\begin{itemize}
	\item Edge tensor with shape $[num\_edges, 3]$ (source, target, timestep)
	\item Node count and per-snapshot node statistics
	\item Feature dimension (768 for BERT)
	\item ID mappings for interpretation
\end{itemize}

\section{Format Data}

\subsection*{Adjacency Matrix Format}

For the GCN operations, adjacency matrices are stored in sparse format:

\begin{equation}
	\hat{A}_t = \tilde{D}_t^{-\frac{1}{2}} \tilde{A}_t \tilde{D}_t^{-\frac{1}{2}}, \quad \tilde{A}_t = A_t + I
\end{equation}

where the normalization enables stable gradient flow through the graph convolution layers.

The sparse representation uses:
\begin{itemize}
	\item \texttt{idx}: $[2, num\_edges]$ tensor of (source, target) pairs
	\item \texttt{vals}: $[num\_edges]$ tensor of edge values (normalized)
\end{itemize}

The number of edges depends on the snapshot considered at the time of training or inference, following the cumulative structure.

% \subsection{Active Node Compaction}

% For incremental learning efficiency, each snapshot uses only the active nodes (those present up to that snapshot). This active node compaction:

% \begin{itemize}
% 	\item Reduces memory usage from $O(N_{total})$ to $O(N_{active})$
% 	\item Remaps adjacency indices to compacted space $[0, N_{active}-1]$
% 	\item Enables training on early snapshots with small node sets
% \end{itemize}

\subsection*{Training Sample Format}

What the model will receive during training for snapshot $t$ is:

\begin{itemize}
	\item \textbf{Historical Adjacency List}: $[\hat{A}_{t-k}, \ldots, \hat{A}_{t-1}]$ for $k$ historical steps
	\item \textbf{Historical Node Features}: $[H_{t-k}, \ldots, H_{t-1}]$ corresponding feature matrices
	\item \textbf{Node Masks}: Binary masks indicating which nodes are active at each historical step
	\item \textbf{Label Edges}: Pairs $(u, v)$ with labels $y \in \{0, 1\}$ for training/evaluation
\end{itemize}

% The negative sampling strategy generates non-existent edges proportional to the positive edges, with smart sampling preferring edges involving nodes that have historical activity.